### 2019-08-11 ch2_dynbenchmark.tex
use zotero instead

Single-cell omics data, including transcriptomics, proteomics and epigenomics data, provide new opportunities for studying cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation \cite{tanay_scaling_2017,Etzrodt2014-tr}. Such dynamic processes can be modeled computationally using trajectory inference (TI) methods, also called pseudotime analysis, which order cells along a trajectory based on similarities in their expression patterns \cite{Trapnell2015,cannoodt_computational_2016,moon_manifoldlearningbasedmethods_2018}. The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also identify more complex trajectory topologies, such as cyclic \cite{liu_reconstructingcellcycle_2017} or disconnected graphs \cite{wolf_graphabstractionreconciles_2017}. TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process1, thereby allowing the objective identification of new (primed) subsets of cells \cite{schlitzer_identificationcdc1cdc2committed_2015}, delineation of a differentiation tree \cite{velten_humanhaematopoieticstem_2017,see_mappinghumandc_2017} and inference of regulatory interactions responsible for one or more bifurcations \cite{Aibar2017a}. Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogs of whole organisms \cite{regev_scienceforumhuman_2017,han_mappingmousecell_2018,schaum_singlecelltranscriptomics20_2018} underline the urgency for accurate, scalable \cite{Aibar2017a,angerer_singlecellsmake_2017} and user-friendly TI methods.

A plethora of TI methods has been developed over the past few years and even more are being created every month (Supplementary Table 1). Indeed, in several repositories listing single-cell tools, such as omictools.org \cite{henry_omictoolsinformativedirectory_2014}, the ‘awesome-single-cell’ list \cite{davis_awesomesinglecelllistsoftware_2018} and scRNA-tools.org \cite{zappia_exploringsinglecellrnaseq_2017}, TI methods are one of the largest categories. While each method has its own unique set of characteristics in terms of underlying algorithm, required prior information and produced outputs, two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory and what type(s) of graph topologies they can detect. Early TI methods typically fixed the topology algorithmically (for example, linear \cite{bendall_singlecelltrajectorydetection_2014,schlitzer_identificationcdc1cdc2committed_2015,shin_singlecellrnaseqwaterfall_2015,campbell_bayesiangaussianprocess_2015} or bifurcating trajectories \cite{Haghverdi2016,Setty2016-og}) or through parameters provided by the user \cite{Trapnell2014,matsumoto_scoupprobabilisticmodel_2016}. These methods therefore mainly focus on correctly ordering the cells along the fixed topology. More recent methods also infer the topology \cite{qiu_reversedgraphembedding_2017,street_slingshotcelllineage_2017,wolf_graphabstractionreconciles_2017}, which increases the difficulty of the problem at hand, but allows the unbiased identification of both the ordering inside a branch and the topology connecting these branches.

Given the diversity in TI methods, it is important to quantitatively assess their performance, scalability, robustness and usability. Many attempts at tackling this issue have already been made \cite{Haghverdi2016,Ji2016-il,Welch2016-ue,matsumoto_scoupprobabilisticmodel_2016,duverle_celltreebioconductorpackage_2016,Cannoodt2016a,lonnberg_singlecellrnaseqcomputational_2017,campbell_probabilisticmodelingbifurcations_2017,wolf_graphabstractionreconciles_2017}, but a comprehensive comparison of TI methods across a large number of different datasets is still lacking. This is problematic, as new users to the field are confronted with an overwhelming choice of TI methods, without a clear idea of which would optimally solve their problem. Moreover, the strengths and weaknesses of existing methods need to be assessed, so that new developments in the field can focus on improving the current state-of-the-art.

@@

We defined several metrics to compare a prediction to a reference trajectory (Supplementary Note 1). Based on an analysis of their robustness and conformity to a set of rules (Supplementary Note 1), we chose four metrics each assessing a different aspect of a trajectory (Figure \ref{fig:figure_1}d): the topology (Hamming–Ipsen–Mikhailov, HIM), the quality of the assignment of cells to branches (F1branches), the cell positions (cordist) and the accuracy of the differentially expressed features along the trajectory (wcorfeatures). The data compendium consisted of both synthetic datasets, which offer the most exact reference trajectory, and real datasets, which provide the highest biological relevance. These real datasets come from a variety of single-cell technologies, organisms and dynamic processes, and contain several types of trajectory topologies (Supplementary Table 2). Real datasets were classified as ‘gold standard’ if the reference trajectory was not extracted from the expression data itself, such as via cellular sorting or cell mixing \cite{tian_scrnaseqmixologybetter_2018}. All other real datasets were classified as ‘silver standard’. For synthetic datasets we used several data simulators, including a simulator of gene regulatory networks using a thermodynamic model of gene regulation \cite{schaffter2011}. For each simulation, we used a real dataset as a reference, to match its dimensions, number of differentially expressed genes, drop-out rates and other statistical properties \cite{zappia_splattersimulationsinglecell_2017}.

@@

While not directly related to the accuracy of the inferred trajectory, it is also important to assess the quality of the implementation and how user-friendly it is for a biological user \cite{taschuk_ten_2017}. We scored each method using a transparent checklist of important scientific and software development practices, including software packaging, documentation, automated code testing and publication into a peer-reviewed journal (Table \ref{tab:scoresheet}). It is important to note that there is a selection bias in the tools chosen for this analysis, as we did not include a substantial set of tools due to issues with installation, code availability and executability on a freely available platform (which excludes MATLAB). The reasons for not including certain tools are all discussed on our repository (https://github.com/dynverse/dynmethods/issues?q=label:unwrappable). Installation issues seem to be quite general in bioinformatics \cite{mangul_comprehensiveanalysisusability_2018} and the trajectory inference field is no exception.

We found that most methods fulfilled the basic criteria, such as the availability of a tutorial and elemental code quality criteria (Figure \ref{fig:figure_3}d and Supplementary Fig. 6). While recent methods had a slightly better quality score than older methods, several quality aspects were consistently lacking for the majority of the methods (Supplementary Fig. 6 right) and we believe that these should receive extra attention from developers. Although these outstanding issues covered all five categories, code assurance and documentation in particular were problematic areas, notwithstanding several studies pinpointing these as good practices \cite{wilson_best_2014,artaza_top10metrics_2016}. Only two methods had a nearly perfect usability score (Slingshot and Celltrails), and these could be used as an inspiration for future methods. We observed no clear relation between usability and method accuracy or usability (Figure \ref{fig:figure_2}b).

@@

		Open source & (1) Method's code is freely available (2) The code can be run on a freely available platform & \cite{l_rpackages_2017,wilson_best_2014,taschuk_ten_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017}\\*

		Version control & The code is available on a public version controlled repository, such as Github & \cite{l_rpackages_2017,wilson_best_2014,taschuk_ten_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017}\\*

		Packaging & (1) The code is provided as a "package", exposing functionality through functions or shell commands (2) The code can be easily installed through a repository such as CRAN, Bioconductor, PyPI, CPAN, debian packages, … & \cite{l_rpackages_2017,wickham_packagesorganizetest_2015,jimenez_foursimplerecommendations_2017,silva_generalguidelinesbiomedical_2017}\\*

		Dependencies & (1) Dependencies are clearly stated in the tutorial or in the code (2) Dependencies are automatically installed & \cite{taschuk_ten_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,karimzadeh_topconsiderationscreating_}\\*

		License & (1) The code is licensed (2) License allows academic use & \cite{l_rpackages_2017,taschuk_ten_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017}\\*

@@

		Function and object naming & (1) Functions/commands have well chosen names (2) Arguments/parameters have well chosen names & \cite{wilson_best_2014,wickham_packagesorganizetest_2015}\\*

		Code style & (1) Code has a consistent style (2) Code follows (basic) good practices in the programming language of choice, for example PEP8 or the tidyverse style guide & \cite{wilson_best_2014,wickham_packagesorganizetest_2015,artaza_top10metrics_2016}\\*

		Code duplication & Duplicated code is minimal & \cite{wilson_best_2014,wickham_packagesorganizetest_2015}\\

		Self-contained functions & The method is exposed to the user as self-contained functions or commands & \cite{anderson_writing_2016,taschuk_ten_2017,silva_generalguidelinesbiomedical_2017}\\*

@@

		Dummy proofing & Package contains dummy proofing, i.e. testing whether the parameters and data supplied by the user make sense and are useful & \cite{l_rpackages_2017,karimzadeh_topconsiderationscreating_}\\

@@

		Unit testing & Method is tested using unit tests & \cite{l_rpackages_2017,wilson_best_2014,anderson_writing_2016,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017}\\*

		Unit testing & Tests are run automatically using functionality from the programming language & \cite{l_rpackages_2017,wilson_best_2014,anderson_writing_2016,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017}\\*

		Continuous integration & The method uses continuous integration, for example on Travis CI & \cite{beaulieu-jones_reproducibility_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017}\\*

@@

		Support & (1) There is a support ticket system, for example on Github (2) The authors respond to tickets and issues are resolved within a reasonable time frame & \cite{wilson_best_2014,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017}\\*

@@

		Function documentation & (1) The purpose and usage of functions/commands is documented (2) The parameters of functions/commands are documented (3) The output of functions/commands is documented & \cite{wilson_best_2014,taschuk_ten_2017,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017,karimzadeh_topconsiderationscreating_}\\*

		Inline documentation & Inline documentation is present in the code & \cite{wilson_best_2014,taschuk_ten_2017,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017,karimzadeh_topconsiderationscreating_}\\*

		Parameter transparency & All important parameters are exposed to the user & \cite{taschuk_ten_2017}\\

@@

		Seed setting & The method does not artificially become deterministic, for example by setting some (0.5) or a lot (1) of seeds & \cite{jfp_greendiceare_2016}\\*

@@

Critical to the broad applicability of TI methods is the standardization of the input and output interfaces of TI methods, so that users can effortlessly execute TI methods on their dataset of interest, compare different predicted trajectories and apply downstream analyses, such as finding genes important for the trajectory, network inference \cite{Aibar2017a} or finding modules of genes \cite{saelens_comprehensiveevaluationmodule_2018}. Our framework is an initial attempt at tackling this problem, and we illustrate its usefulness here by comparing the predicted trajectories of several top-performing methods on datasets containing a linear, tree, cyclic and disconnected graph topology (Figure \ref{fig:figure_6}). Using our framework, this figure can be recreated using only a couple of lines of R code (https://methods.dynverse.org). In the future, this framework could be extended to allow additional input data, such as spatial and RNA velocity information \cite{manno_rnavelocitysingle_2018}, and easier downstream analyses. In addition, further discussion within the field is required to arrive at a consensus concerning a common interface for trajectory models, which can include additional features such as uncertainty and gene importance.

@@

We found that the performance of a method can be very variable between datasets, and therefore included a large set of both real and synthetic data within our evaluation, leading to a robust overall ranking of the different methods. However, ‘good-yet-not-the-best’ methods \cite{Norel2011} can still provide a very valuable contribution to the field, especially if they make use of novel algorithms, return a more scalable solution or provide a unique insight in specific use cases. This is also supported by our analysis of method complementarity. Some examples for the latter include PhenoPath, which can include additional covariates in its model, ouija, which returns a measure of uncertainty of each cell’s position within the trajectory, and StemID, which can infer the directionality of edges within the trajectory.

@@

We gathered a list of 71 trajectory inference tools (Supplementary Table 1) by searching the literature for ‘trajectory inference’ and ‘pseudotemporal ordering’, and based on two existing lists found online: \href{https://github.com/seandavi/awesome-single-cell}{https://github.com/seandavi/awesome-single-cell} \cite{davis_awesomesinglecelllistsoftware_2018} and \href{https://github.com/agitter/single-cell-pseudotime}{https://github.com/agitter/single-cell-pseudotime} \cite{gitter_singlecellpseudotimeoverviewalgorithms_2018}. We welcome any contributions by creating an issue at https://methods.dynverse.org.

@@

	\item \textbf{Component 1}: This method returns the first component of a principal component analysis (PCA) dimensionality reduction as a linear trajectory. This method is especially relevant as it has been used in a few studies already \cite{Kouno2013-km,zeng_pseudotemporalorderingsingle_2017}.

@@

\begin{itemize}
	\item \textbf{dyngen}: simulations of gene regulatory networks, available at \href{https://github.com/dynverse/dyngen}{https://github.com/dynverse/dyngen}
	\item \textbf{dyntoy}: random gradients of expression in the reduced space, available at \href{https://github.com/dynverse/dyntoy}{https://github.com/dynverse/dyntoy}
	\item \textbf{PROSSTT}: expression is sampled from a linear model that depends on pseudotime \cite{papadopoulosPROSSTTProbabilisticSimulation2018}
	\item \textbf{Splatter}: simulations of non-linear paths between different expression states \cite{zappia_splattersimulationsinglecell_2017}
\end{itemize}

@@

The dyngen (\href{https://github.com/dynverse/dyngen}{https://github.com/dynverse/dyngen}) workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods \cite{schaffter2011,marbach_wisdom_2012} and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment. At every step, we tried to mirror real regulatory networks, while keeping the model simple and easily extendable. We simulated a total of 110 datasets, with 11 different topologies.

@@

One of the main processes involved in cellular dynamic processes is gene regulation, where regulatory cascades and feedback loops lead to progressive changes in expression and decision making. The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested in vivo. One driver of bifurcation seems to be mutual antagonism, where genes \cite{Xu2015-kn} strongly repress each other, forcing one of the two to become inactive \cite{graf_forcingcellschange_2009}. Such mutual antagonism can be modelled and simulated \cite{wang_quantifyingwaddingtonlandscape_2011,ferrell_bistabilitybifurcationswaddington_2012}. Although such a two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into modules) repressing each other \cite{Yosef2013-mj}.

@@

To simulate the gene regulatory network, we used a system of differential equations similar to those used in evaluations of gene regulatory network inference methods \cite{marbach_wisdom_2012}. In this model, the changes in gene expression ($x_i$) and protein expression ($y_i$) are modeled using ordinary differential equations \cite{schaffter2011} (ODEs):

@@

Similar to GeneNetWeaver \cite{schaffter2011}, we sample the different parameters from random distributions, defined as follows. $e$ defines whether a transcription factor activates (1) or represses (-1), as defined within the regulatory network network.

@@

PROSSTT is a recent data simulator \cite{papadopoulosPROSSTTProbabilisticSimulation2018}, which simulates expression using linear mixtures of expression programs and random walks through the trajectory. We used 5 topology generators from dyntoy (linear, bifurcating, multifurcating, binary tree and tree), and simulated for each topology generator 10 datasets using different reference real datasets. However, due to frequent crashes of the tool, only 19 datasets created output and were thus used in the evaluation.

@@

The importance of using multiple metrics to compare complex models has been stated repeatedly \cite{Norel2011}. Furthermore, a trajectory is a model with multiple layers of complexity, which calls for several metrics each assessing a different layer. We therefore defined several possible metrics for comparing trajectories, each investigating different layers. These are all discussed in Supplementary Note 1 along with examples and robustness analyses when appropriate.

@@

The HIM metric \cite{jurman_HIM_Glocal_Metric_2015} uses the two weighted adjacency matrices of the milestone networks as input (weighted by edge length). It is a linear combination of the normalized Hamming distance, which gives an indication of the differences in edge lengths, and the normalized Ipsen–Mikhailov distance, which assesses the similarity in degree distributions. The latter has a parameter $\gamma$, which was fixed at 0.1 to make the scores comparable between datasets. We illustrate the metric and discuss alternatives in Supplementary Note 1.

@@

The \textit{HIM} metric (Hamming-Ipsen-Mikhailov distance) \cite{jurman_HIM_Glocal_Metric_2015} which was adopted from the R nettools package (\href{https://github.com/filosi/nettools}{https://github.com/filosi/nettools}). It uses an adjacency matrix which was weighted according to the lengths of each edges within the milestone network. Conceptually, \textit{HIM} is a linear combination of:

>>>>

Single-cell omics data, including transcriptomics, proteomics and epigenomics data, provide new opportunities for studying cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation \cite{tanay_scalingsinglecellgenomics_2017,etzrodt_quantitativesinglecellapproaches_2014}. Such dynamic processes can be modeled computationally using trajectory inference (TI) methods, also called pseudotime analysis, which order cells along a trajectory based on similarities in their expression patterns \cite{trapnell_definingcelltypes_2015,cannoodt_computationalmethodstrajectory_2016,moon_manifoldlearningbasedmethods_2018}. The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also identify more complex trajectory topologies, such as cyclic \cite{liu_reconstructingcellcycle_2017} or disconnected graphs \cite{wolf_graphabstractionreconciles_2017}. TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process\cite{tanay_scalingsinglecellgenomics_2017}, thereby allowing the objective identification of new (primed) subsets of cells \cite{schlitzer_identificationcdc1cdc2committed_2015}, delineation of a differentiation tree \cite{velten_humanhaematopoieticstem_2017,see_mappinghumandc_2017} and inference of regulatory interactions responsible for one or more bifurcations \cite{aibar_scenicsinglecellregulatory_2017}. Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogs of whole organisms \cite{regev_humancellatlas_2017,han_mappingmousecell_2018,schaum_singlecelltranscriptomics20_2018} underline the urgency for accurate, scalable \cite{aibar_scenicsinglecellregulatory_2017,angerer_singlecellsmake_2017} and user-friendly TI methods.

A plethora of TI methods has been developed over the past few years and even more are being created every month (Supplementary Table 1). Indeed, in several repositories listing single-cell tools, such as omictools.org \cite{henry_omictoolsinformativedirectory_2014}, the ‘awesome-single-cell’ list \cite{davis_awesomesinglecell_2018} and scRNA-tools.org \cite{zappia_exploringsinglecellrnaseq_2017}, TI methods are one of the largest categories. While each method has its own unique set of characteristics in terms of underlying algorithm, required prior information and produced outputs, two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory and what type(s) of graph topologies they can detect. Early TI methods typically fixed the topology algorithmically (for example, linear \cite{bendall_singlecelltrajectorydetection_2014,schlitzer_identificationcdc1cdc2committed_2015,shin_singlecellrnaseqwaterfall_2015,campbell_bayesiangaussianprocess_2015} or bifurcating trajectories \cite{haghverdi_diffusionpseudotimerobustly_2016,setty_wishboneidentifiesbifurcating_2016}) or through parameters provided by the user \cite{trapnell_dynamicsregulatorscell_2014,matsumoto_scoupprobabilisticmodel_2016}. These methods therefore mainly focus on correctly ordering the cells along the fixed topology. More recent methods also infer the topology \cite{qiu_reversedgraphembedding_2017,street_slingshotcelllineage_2018,wolf_graphabstractionreconciles_2017}, which increases the difficulty of the problem at hand, but allows the unbiased identification of both the ordering inside a branch and the topology connecting these branches.

Given the diversity in TI methods, it is important to quantitatively assess their performance, scalability, robustness and usability. Many attempts at tackling this issue have already been made \cite{haghverdi_diffusionpseudotimerobustly_2016,ji_tscanpseudotimereconstruction_2016,welch_slicerinferringbranched_2016,matsumoto_scoupprobabilisticmodel_2016,duverle_celltreebioconductorpackage_2016,cannoodt_computationalmethodstrajectory_2016,lonnberg_singlecellrnaseqcomputational_2017,campbell_probabilisticmodelingbifurcations_2017,wolf_graphabstractionreconciles_2017}, but a comprehensive comparison of TI methods across a large number of different datasets is still lacking. This is problematic, as new users to the field are confronted with an overwhelming choice of TI methods, without a clear idea of which would optimally solve their problem. Moreover, the strengths and weaknesses of existing methods need to be assessed, so that new developments in the field can focus on improving the current state-of-the-art.

@@

We defined several metrics to compare a prediction to a reference trajectory (Supplementary Note 1). Based on an analysis of their robustness and conformity to a set of rules (Supplementary Note 1), we chose four metrics each assessing a different aspect of a trajectory (Figure \ref{fig:figure_1}d): the topology (Hamming–Ipsen–Mikhailov, HIM), the quality of the assignment of cells to branches (F1branches), the cell positions (cordist) and the accuracy of the differentially expressed features along the trajectory (wcorfeatures). The data compendium consisted of both synthetic datasets, which offer the most exact reference trajectory, and real datasets, which provide the highest biological relevance. These real datasets come from a variety of single-cell technologies, organisms and dynamic processes, and contain several types of trajectory topologies (Supplementary Table 2). Real datasets were classified as ‘gold standard’ if the reference trajectory was not extracted from the expression data itself, such as via cellular sorting or cell mixing \cite{tian_scrnaseqmixologybetter_2018}. All other real datasets were classified as ‘silver standard’. For synthetic datasets we used several data simulators, including a simulator of gene regulatory networks using a thermodynamic model of gene regulation \cite{schaffter_genenetweaversilicobenchmark_2011}. For each simulation, we used a real dataset as a reference, to match its dimensions, number of differentially expressed genes, drop-out rates and other statistical properties \cite{zappia_splattersimulationsinglecell_2017}.

@@

While not directly related to the accuracy of the inferred trajectory, it is also important to assess the quality of the implementation and how user-friendly it is for a biological user \cite{taschuk_tensimplerules_2017}. We scored each method using a transparent checklist of important scientific and software development practices, including software packaging, documentation, automated code testing and publication into a peer-reviewed journal (Table \ref{tab:scoresheet}). It is important to note that there is a selection bias in the tools chosen for this analysis, as we did not include a substantial set of tools due to issues with installation, code availability and executability on a freely available platform (which excludes MATLAB). The reasons for not including certain tools are all discussed on our repository (https://github.com/dynverse/dynmethods/issues?q=label:unwrappable). Installation issues seem to be quite general in bioinformatics \cite{mangul_comprehensiveanalysisusability_2018} and the trajectory inference field is no exception.

We found that most methods fulfilled the basic criteria, such as the availability of a tutorial and elemental code quality criteria (Figure \ref{fig:figure_3}d and Supplementary Fig. 6). While recent methods had a slightly better quality score than older methods, several quality aspects were consistently lacking for the majority of the methods (Supplementary Fig. 6 right) and we believe that these should receive extra attention from developers. Although these outstanding issues covered all five categories, code assurance and documentation in particular were problematic areas, notwithstanding several studies pinpointing these as good practices \cite{wilson_bestpracticesscientific_2014,artaza_top10metrics_2016}. Only two methods had a nearly perfect usability score (Slingshot and Celltrails), and these could be used as an inspiration for future methods. We observed no clear relation between usability and method accuracy or usability (Figure \ref{fig:figure_2}b).

@@

		Open source & (1) Method's code is freely available (2) The code can be run on a freely available platform & \cite{lee_rpackagespackagedevelopment_2017,wilson_bestpracticesscientific_2014,taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017}\\*

		Version control & The code is available on a public version controlled repository, such as Github & \cite{lee_rpackagespackagedevelopment_2017,wilson_bestpracticesscientific_2014,taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017}\\*

		Packaging & (1) The code is provided as a "package", exposing functionality through functions or shell commands (2) The code can be easily installed through a repository such as CRAN, Bioconductor, PyPI, CPAN, debian packages, … & \cite{lee_rpackagespackagedevelopment_2017,wickham_packagesorganizetest_2015,jimenez_foursimplerecommendations_2017,silva_generalguidelinesbiomedical_2017}\\*

		Dependencies & (1) Dependencies are clearly stated in the tutorial or in the code (2) Dependencies are automatically installed & \cite{taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,karimzadeh_topconsiderationscreating_}\\*

		License & (1) The code is licensed (2) License allows academic use & \cite{lee_rpackagespackagedevelopment_2017,taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017}\\*

@@

		Function and object naming & (1) Functions/commands have well chosen names (2) Arguments/parameters have well chosen names & \cite{wilson_bestpracticesscientific_2014,wickham_packagesorganizetest_2015}\\*

		Code style & (1) Code has a consistent style (2) Code follows (basic) good practices in the programming language of choice, for example PEP8 or the tidyverse style guide & \cite{wilson_bestpracticesscientific_2014,wickham_packagesorganizetest_2015,artaza_top10metrics_2016}\\*

		Code duplication & Duplicated code is minimal & \cite{wilson_bestpracticesscientific_2014,wickham_packagesorganizetest_2015}\\

		Self-contained functions & The method is exposed to the user as self-contained functions or commands & \cite{anderson_writinggreatscientific_2016,taschuk_tensimplerules_2017,silva_generalguidelinesbiomedical_2017}\\*

@@

		Dummy proofing & Package contains dummy proofing, i.e. testing whether the parameters and data supplied by the user make sense and are useful & \cite{lee_rpackagespackagedevelopment_2017,karimzadeh_topconsiderationscreating_}\\

@@

		Unit testing & Method is tested using unit tests & \cite{lee_rpackagespackagedevelopment_2017,wilson_bestpracticesscientific_2014,anderson_writinggreatscientific_2016,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017}\\*

		Unit testing & Tests are run automatically using functionality from the programming language & \cite{lee_rpackagespackagedevelopment_2017,wilson_bestpracticesscientific_2014,anderson_writinggreatscientific_2016,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017}\\*

		Continuous integration & The method uses continuous integration, for example on Travis CI & \cite{beaulieu-jones_reproducibilitycomputationalworkflows_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017}\\*

@@

		Support & (1) There is a support ticket system, for example on Github (2) The authors respond to tickets and issues are resolved within a reasonable time frame & \cite{wilson_bestpracticesscientific_2014,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017}\\*

@@

		Function documentation & (1) The purpose and usage of functions/commands is documented (2) The parameters of functions/commands are documented (3) The output of functions/commands is documented & \cite{wilson_bestpracticesscientific_2014,taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017,karimzadeh_topconsiderationscreating_}\\*

		Inline documentation & Inline documentation is present in the code & \cite{wilson_bestpracticesscientific_2014,taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017,karimzadeh_topconsiderationscreating_}\\*

		Parameter transparency & All important parameters are exposed to the user & \cite{taschuk_tensimplerules_2017}\\

@@

		Seed setting & The method does not artificially become deterministic, for example by setting some (0.5) or a lot (1) of seeds & \cite{puget_greendiceare_2016}\\*

@@

Critical to the broad applicability of TI methods is the standardization of the input and output interfaces of TI methods, so that users can effortlessly execute TI methods on their dataset of interest, compare different predicted trajectories and apply downstream analyses, such as finding genes important for the trajectory, network inference \cite{aibar_scenicsinglecellregulatory_2017} or finding modules of genes \cite{saelens_comprehensiveevaluationmodule_2018}. Our framework is an initial attempt at tackling this problem, and we illustrate its usefulness here by comparing the predicted trajectories of several top-performing methods on datasets containing a linear, tree, cyclic and disconnected graph topology (Figure \ref{fig:figure_6}). Using our framework, this figure can be recreated using only a couple of lines of R code (https://methods.dynverse.org). In the future, this framework could be extended to allow additional input data, such as spatial and RNA velocity information \cite{manno_rnavelocitysingle_2018}, and easier downstream analyses. In addition, further discussion within the field is required to arrive at a consensus concerning a common interface for trajectory models, which can include additional features such as uncertainty and gene importance.

@@

We found that the performance of a method can be very variable between datasets, and therefore included a large set of both real and synthetic data within our evaluation, leading to a robust overall ranking of the different methods. However, ‘good-yet-not-the-best’ methods \cite{norel_selfassessmenttrapcan_2011} can still provide a very valuable contribution to the field, especially if they make use of novel algorithms, return a more scalable solution or provide a unique insight in specific use cases. This is also supported by our analysis of method complementarity. Some examples for the latter include PhenoPath, which can include additional covariates in its model, ouija, which returns a measure of uncertainty of each cell’s position within the trajectory, and StemID, which can infer the directionality of edges within the trajectory.

@@

We gathered a list of 71 trajectory inference tools (Supplementary Table 1) by searching the literature for ‘trajectory inference’ and ‘pseudotemporal ordering’, and based on two existing lists found online: \href{https://github.com/seandavi/awesome-single-cell}{https://github.com/seandavi/awesome-single-cell} \cite{davis_awesomesinglecell_2018} and \href{https://github.com/agitter/single-cell-pseudotime}{https://github.com/agitter/single-cell-pseudotime} \cite{gitter_singlecellrnaseqpseudotime_2018}. We welcome any contributions by creating an issue at https://methods.dynverse.org.

@@

	\item \textbf{Component 1}: This method returns the first component of a principal component analysis (PCA) dimensionality reduction as a linear trajectory. This method is especially relevant as it has been used in a few studies already \cite{kouno_temporaldynamicstranscriptional_2013,zeng_pseudotemporalorderingsingle_2017}.

@@

\begin{itemize}
	\item \textbf{dyngen}: simulations of gene regulatory networks, available at \href{https://github.com/dynverse/dyngen}{https://github.com/dynverse/dyngen}
	\item \textbf{dyntoy}: random gradients of expression in the reduced space, available at \href{https://github.com/dynverse/dyntoy}{https://github.com/dynverse/dyntoy}
	\item \textbf{PROSSTT}: expression is sampled from a linear model that depends on pseudotime \cite{papadopoulos_prossttprobabilisticsimulation_2018}
	\item \textbf{Splatter}: simulations of non-linear paths between different expression states \cite{zappia_splattersimulationsinglecell_2017}
\end{itemize}

@@

The dyngen (\href{https://github.com/dynverse/dyngen}{https://github.com/dynverse/dyngen}) workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods \cite{schaffter_genenetweaversilicobenchmark_2011,marbach_wisdomcrowdsrobust_2012} and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment. At every step, we tried to mirror real regulatory networks, while keeping the model simple and easily extendable. We simulated a total of 110 datasets, with 11 different topologies.

@@

One of the main processes involved in cellular dynamic processes is gene regulation, where regulatory cascades and feedback loops lead to progressive changes in expression and decision making. The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested in vivo. One driver of bifurcation seems to be mutual antagonism, where genes \cite{xu_regulationbifurcatingcell_2015} strongly repress each other, forcing one of the two to become inactive \cite{graf_forcingcellschange_2009}. Such mutual antagonism can be modelled and simulated \cite{wang_quantifyingwaddingtonlandscape_2011,ferrell_bistabilitybifurcationswaddington_2012}. Although such a two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into modules) repressing each other \cite{yosef_dynamicregulatorynetwork_2013}.

@@

To simulate the gene regulatory network, we used a system of differential equations similar to those used in evaluations of gene regulatory network inference methods \cite{marbach_wisdomcrowdsrobust_2012}. In this model, the changes in gene expression ($x_i$) and protein expression ($y_i$) are modeled using ordinary differential equations \cite{schaffter_genenetweaversilicobenchmark_2011} (ODEs):

@@

Similar to GeneNetWeaver \cite{schaffter_genenetweaversilicobenchmark_2011}, we sample the different parameters from random distributions, defined as follows. $e$ defines whether a transcription factor activates (1) or represses (-1), as defined within the regulatory network network.

@@

PROSSTT is a recent data simulator \cite{papadopoulos_prossttprobabilisticsimulation_2018}, which simulates expression using linear mixtures of expression programs and random walks through the trajectory. We used 5 topology generators from dyntoy (linear, bifurcating, multifurcating, binary tree and tree), and simulated for each topology generator 10 datasets using different reference real datasets. However, due to frequent crashes of the tool, only 19 datasets created output and were thus used in the evaluation.

@@

The importance of using multiple metrics to compare complex models has been stated repeatedly \cite{norel_selfassessmenttrapcan_2011}. Furthermore, a trajectory is a model with multiple layers of complexity, which calls for several metrics each assessing a different layer. We therefore defined several possible metrics for comparing trajectories, each investigating different layers. These are all discussed in Supplementary Note 1 along with examples and robustness analyses when appropriate.

@@

The HIM metric \cite{jurman_himglocalmetric_2015} uses the two weighted adjacency matrices of the milestone networks as input (weighted by edge length). It is a linear combination of the normalized Hamming distance, which gives an indication of the differences in edge lengths, and the normalized Ipsen–Mikhailov distance, which assesses the similarity in degree distributions. The latter has a parameter $\gamma$, which was fixed at 0.1 to make the scores comparable between datasets. We illustrate the metric and discuss alternatives in Supplementary Note 1.

@@

The \textit{HIM} metric (Hamming-Ipsen-Mikhailov distance) \cite{jurman_himglocalmetric_2015} which was adopted from the R nettools package (\href{https://github.com/filosi/nettools}{https://github.com/filosi/nettools}). It uses an adjacency matrix which was weighted according to the lengths of each edges within the milestone network. Conceptually, \textit{HIM} is a linear combination of:
